NumLin: Linear Types for Linear Algebra

### 2019-04-12 write-up/paper/discussion.tex
Fix formatting

The original implementation used 6 temporaries, one of which was immediately
spotted as never being used due to linearity. It also contained two variables
which were marked as \texttt{intent(in)} but would have been written over by
calls to `gemm', spotted by the fractional-capabilities feature. Furthermore,
it used a matrix \emph{twice} in a call to `symm', once with a read permission
but once with a \emph{write} permission.  Fortran assumes that any parameter
being written to is not aliased and so this call was not only incorrect, but
illegal according to the standard, both aspects of which were captured by
linearity and fractional-capabilities.

@@

Using linear types for BLAS routines is a particularly good domain fit
(given the implicit restrictions on aliasing arguments), and as a
result the idea of using substructural types to express array
computations is not a particularly new
one\cite{scholz,henriksen,bernardy2016}.  However, many of these
designs have been focused on building languages to \emph{implement}
the kernel linear algebra functions, and as a result, they tend to add
additional limitations on the language design. Both
Futhark~\cite{henriksen} and Single Assignment C~\cite{scholz} omit
higher-order functions to facilitate compilation to GPUs. The work of
\cite{bernardy2016} forbids term-level recursion, in order to ensure
that all higher-order computations can be statically normalized away
and thereby maximize opportunities for array fusion.

@@

This approach is similar to the one taken in linear algebra libraries
for Rust -- these libraries typically take advantage of the
distinction that Rust's type system offers between mutable
views/references to arrays.  The work of \cite{weiss} and
\cite{rustbelt} suggest that Rust's borrow-checker can be expressed
in simpler terms using \emph{fractional-permissions}, though to our
knowledge the programmer-visible lifetime analysis in Rust has never
been formalized.

@@

Going forwards, one may wish to eliminate even more runtime errors from \lang,
by extending its type system. For example, we could have used existential types
to statically track pointer identities\cite{morrisett}, or parametric
polymorphism.

We could also attempt to catch mismatched dimensions at compile time
as well.  While this could be done with generative phantom
types\cite{abe2015simple}, using dependent types may offer more
flexibility in \emph{partitioning} regions\cite{space_monads} or
statically enforcing dimensions related constraints of the arguments
at compile-time.  ATS\cite{cui2005ats} is an example of a language
which combines linear types with a sophisticated proof layer. But
although it provides BLAS bindings, it does not aim to provide
aliasing restrictions as demonstrated in this paper.

Taking this idea one step even further, since matrix dimensions are typically
fixed at runtime, we could \emph{stage} \lang\ programs and compile matrix
expressions using more sophisticated algorithms\cite{barthels}. However, it is
worth noting that without care, such algorithms\cite{rocklin_thesis}, usually
based on graph-based, ad-hoc dataflow analysis, can produce erroneous output
which would not get past a linear type system with fractions.

We also think that this concept (and the general design of its implementation)
need not be limited to linear algebra: we could conceivably `backport' this
idea to other contexts that need linearity (concurrency, single-use
continuations, zero-copy buffer, streaming I/O) or combine it with dependent
types to achieve even more expressive power to split up a single block of
memory into multiple regions in an arbitrary manner\cite{space_monads}.

>>>>

The original implementation used 6 temporaries, one of which was immediately
spotted as never being used due to linearity. It also contained two variables
which were marked as \texttt{intent(in)} but would have been written over by
calls to `gemm', spotted by the fractional capabilities feature. Furthermore,
it used a matrix \emph{twice} in a call to `symm', once with a read permission
but once with a \emph{write} permission.  Fortran assumes that any parameter
being written to is not aliased and so this call was not only incorrect, but
illegal according to the standard, both aspects of which were captured by
linearity and fractional capabilities.

@@

Using linear types for BLAS routines is a particularly good domain fit (given
the implicit restrictions on aliasing arguments), and as a result the idea of
using substructural types to express array computations is not a particularly
new one~\cite{scholz,henriksen,bernardy2016}.  However, many of these designs
have been focused on building languages to \emph{implement} the kernel linear
algebra functions, and as a result, they tend to add additional limitations on
the language design. Both Futhark~\cite{henriksen} and Single Assignment
C~\cite{scholz} omit higher-order functions to facilitate compilation to GPUs.
The work of \cite{bernardy2016} forbids term-level recursion, in order to
ensure that all higher-order computations can be statically normalized away and
thereby maximize opportunities for array fusion.

@@

This approach is similar to the one taken in linear algebra libraries for Rust
-- these libraries typically take advantage of the distinction that Rust's type
system offers between mutable views/references to arrays.  The work of
\cite{weiss} and \cite{rustbelt} suggest that Rust's borrow-checker \emph{can
be expressed in simpler terms} using fractional permissions, though to our
knowledge the programmer-visible lifetime analysis in Rust has never been
formalized.

@@

Going forwards, one may wish to eliminate even more runtime errors from \lang,
by extending its type system. For example, we could have used existential types
to statically track pointer identities~\cite{morrisett}, or parametric
polymorphism.

We could also attempt to catch mismatched dimensions at compile time as well.
While this could be done with generative phantom types~\cite{abe2015simple},
using dependent types may offer more flexibility in \emph{partitioning}
regions~\cite{space_monads} or statically enforcing dimensions related
constraints of the arguments at compile-time.  ATS~\cite{cui2005ats} is an
example of a language which combines linear types with a sophisticated proof
layer. But although it provides BLAS bindings, it does not aim to provide
aliasing restrictions as demonstrated in this paper.

Taking this idea one step even further, since matrix dimensions are typically
fixed at runtime, we could \emph{stage} \lang\ programs and compile matrix
expressions using more sophisticated algorithms~\cite{barthels}. However, it is
worth noting that without care, such algorithms~\cite{rocklin_thesis}, usually
based on graph-based, ad-hoc dataflow analysis, can produce erroneous output
which would not get past a linear type system with fractions.

We also think that this concept (and the general design of its implementation)
need not be limited to linear algebra: we could conceivably `backport' this
idea to other contexts that need linearity (concurrency, single-use
continuations, zero-copy buffer, streaming I/O) or combine it with dependent
types to achieve even more expressive power to split up a single block of
memory into multiple regions in an arbitrary manner~\cite{space_monads}.

### 2019-04-12 write-up/paper/formal_system.tex
Fix formatting

Using this, we can explain how the ! annotation on variables -- first
introduced in the factorial example in \ref{subsubsec:factorial} -- works. That
is, we can explain why the meaning of $\mathbf{let}\ !x = e\ \mathbf{in}\ e'$
can be expressed using only the rules presented thus far, as $\mathbf{let\
Many}\ x = e\ \mathbf{in\ } \mathbf{let\ Many}\ x = \mathbf{Many}\
(\mathbf{Many}\ x)\ \mathbf{in}\ e'$.
\footnote{\emph{Why} we have this at all is for the sake of ergonomics when
    using binary arithmetic operations (for example, of type
    $!\ottkw{int} \, \otimes \, !\ottkw{int} \multimap \; !\ottkw{int}$): a
    programmer should be able to write \highl{let x = 5 + 5 in x - x},
    which, although non-linear in \highl{x}, is morally right because
    integers and operations on them rarely need to be linear. Though it
    should be possible to handle this using a LNL-style presentation of
    linear types~\cite{benton1995} (using adjoint modalities to distinguish
    between intrinsically linear and intrinsically intuitionistic types) that
    is a pretty big digression from the stated goals of this paper.}
The reader is invited to quickly convince themselves that the following
meta-rule is provable using \textsc{Ty\_Bang\_Intro} (twice),
\textsc{Ty\_Bang\_Elim} (twice) and weakening the intuitionistic environment
$\Delta$ (once).

\vspace{-\baselineskip}
\[
\ottdrule{%
\ottpremise{\Theta  \ottsym{;}  \Delta  \ottsym{;}  \Gamma  \vdash  \ottnt{e}  \ottsym{:}  \: !  \ottnt{t}}%
\ottpremise{\Theta  \ottsym{;}  \Delta  \ottsym{,}  \mathit{x}  \ottsym{:}  \: !  \ottnt{t}  \ottsym{;}  \Gamma'  \vdash  \ottnt{e'}  \ottsym{:}  \ottnt{t'}}%
}{
\Theta  \ottsym{;}  \Delta  \ottsym{;}  \Gamma  \ottsym{,}  \Gamma'  \vdash  \ottkw{let} \, ! \mathit{x} \ottsym{=} \ottnt{e} \, \ottkw{in} \, \ottnt{e'}  \ottsym{:}  \ottnt{t'}}{%
{\ottdrulename{Meta\_Ty\_Bang\_Var\_Elim}}{}%
}
\]

@@

because after the $i$ steps, we need to invoke \textsc{Op\_Let\_Var} to proceed
evalution for any remaining $j-i$ steps. After that, it suffices to use the
induction hypothesis on the second sub-expression to finish the proof.  To do
so, we need to construct a valid linear substitution and heap (i.e., one in
$\den{L}{k}{\Gamma', x : t}\theta$). We take the other `half' of the linear
substitution and heap (from the inital split at the start) and extend it with
$[x \mapsto v]$, (where $x$ is the variable bound in the let-expression and $v$
is the value we assume the first sub-expression evaluated to in $i$ steps).

>>>>

Using this, we can explain how the ! annotation on variables -- first
introduced in the factorial example in \ref{subsubsec:factorial} -- works. That
is, we can explain why the meaning of $\mathbf{let}\ !x = e\ \mathbf{in}\ e'$
can be expressed using only the rules presented thus far, as $\mathbf{let\
Many}\ x = e\ \mathbf{in\ } \mathbf{let\ Many}\ x = \mathbf{Many}\
(\mathbf{Many}\ x)\ \mathbf{in}\ e'$.\footnote{\emph{Why} we have this at all
    is for the sake of ergonomics when using binary arithmetic operations
    (e.g.\ of type $!\ottkw{int} \, \otimes \, !\ottkw{int} \multimap \;
    !\ottkw{int}$): a programmer should be able to write \highl{let x = 5 + 5
    in x - x}, which, although non-linear in \highl{x}, is morally right
    because integers and operations on them rarely need to be linear. Though it
    should be possible to handle this using a LNL-style presentation of linear
    types~\cite{benton1995} (using adjoint modalities to distinguish between
    intrinsically linear and intrinsically intuitionistic types) that is a
    pretty big digression from the stated goals of this paper.}
The reader is invited to quickly convince themselves that the following
meta-rule is provable using \textsc{Ty\_Bang\_Intro} (twice),
\textsc{Ty\_Bang\_Elim} (twice) and weakening the intuitionistic environment
$\Delta$ (once).

\vspace{-\baselineskip}
\[
\ottdrule{%
\ottpremise{\Theta  \ottsym{;}  \Delta  \ottsym{;}  \Gamma  \vdash  \ottnt{e}  \ottsym{:}  \: !  \ottnt{t}}%
\ottpremise{\Theta  \ottsym{;}  \Delta  \ottsym{,}  \mathit{x}  \ottsym{:}  \: !  \ottnt{t}  \ottsym{;}  \Gamma'  \vdash  \ottnt{e'}  \ottsym{:}  \ottnt{t'}}%
}{
\Theta  \ottsym{;}  \Delta  \ottsym{;}  \Gamma  \ottsym{,}  \Gamma'  \vdash  \ottkw{let} \, ! \mathit{x} \ottsym{=} \ottnt{e} \, \ottkw{in} \, \ottnt{e'}  \ottsym{:}  \ottnt{t'}}{%
{\ottdrulename{Meta\_Ty\_Let\_Bang}}{}%
}
\]

@@

because after the $i$ steps, we need to invoke \textsc{Op\_Let\_Var} to proceed
evalution for any remaining $j-i$ steps. After that, it suffices to use the
induction hypothesis on the second sub-expression to finish the proof.  To do
so, we need to construct a valid linear substitution and heap (one in
$\den{L}{k}{\Gamma', x : t}\theta$). We take the other `half' of the linear
substitution and heap (from the inital split at the start) and extend it with
$[x \mapsto v]$, (where $x$ is the variable bound in the let-expression and $v$
is the value we assume the first sub-expression evaluated to in $i$ steps).

### 2019-04-12 write-up/paper/implementation.tex
Fix formatting

Desugaring is conventional, outlined in Appendix~\ref{sec:lang_desugar}.  Matrix
expression are translated into BLAS/ LAPACK calls via purely syntactic
pattern-matching, outlined in Figure~\ref{fig:lang_matexp}.

@@

\begin{figure}[t]
\begin{center}
\[\def\arraystretch{1.3}
    \begin{array}{rcl}
    \mathbf{let\ } v \leftarrow x[e]\ \mathbf{in\ } e &
    \Rightarrow &
    \mathbf{let\ } (x, !v)\ = x[e]\ \mathbf{in\ } e \qquad \textrm{(similarly for matrices)}
\\
    \mathbf{let\ } x_2 \leftarrow \mathbf{new}\ [|\ x_1\ |]\ \mathbf{in\ } e &
    \Rightarrow &
    \mathbf{let\ } (x_1, x_2)\ = \mathbf{copyM\ } \_\ x_1\ \mathbf{in\ } e
\\
    \mathbf{let\ } x_2 \leftarrow [|\ x_1\ |]\ \mathbf{in\ } e &
    \Rightarrow &
    \mathbf{let\ } (x_1, x_2)\ = \mathbf{copyM\_to\ } \_\  x_1\ x_2\ \mathbf{in\ } e
    \end{array}
\]
\[
    M ::= X\ \mid\ X^T\ \mid\ \textrm{sym}( X )
\]
\end{center}
\begin{align*}
% new
    \mathbf{let\ } &Y \leftarrow \mathbf{new}\ (n, k)\ [|\ \alpha M_1 M_2 \ |]\ \mathbf{in\ } e
    \Rightarrow \\
    & \mathbf{let\ } Y\ = \mathbf{matrix\ } n\ k\ \mathbf{in}
    \ \mathbf{let\ } Y \leftarrow [|\ \alpha M_1 M_2 + 0Y\ |]\ \mathbf{in\ } e
\\[0.1\baselineskip]
%syrk
    \mathbf{let\ } &Y \leftarrow [|\ \alpha X X^T + \beta Y\ |]\ \mathbf{in\ } e
    \Rightarrow \\
    & \mathbf{let\ } (X,Y)\ = \mathbf{syrk\ } \mathbf{false}
    \ \alpha\ \_\ X\ \beta\ Y\ \mathbf{in\ } e
\\[0.1\baselineskip]
    \mathbf{let\ } &Y \leftarrow [|\ \alpha X^T X + \beta Y\ |]\ \mathbf{in\ } e
    \Rightarrow \\
    & \mathbf{let\ } (X,Y)\ = \mathbf{syrk\ } \mathbf{true}
    \ \alpha\ \_\ X\ \beta\ Y\ \mathbf{in\ } e
%symm
\\[0.1\baselineskip]
    \mathbf{let\ } &Y \leftarrow\ [|\ \alpha\,\textrm{sym}(X_1)\,X_2 + \beta Y\ |]\ \mathbf{in\ } e
    \Rightarrow \\
    & \mathbf{let\ } ((X_1,X_2),Y)\ = \mathbf{symm\ } \mathbf{false}
        \ \alpha\ \_ \ X_1 \_\ X_2\ \beta\ Y\ \mathbf{in\ } e
\\[0.1\baselineskip]
    \mathbf{let\ } &Y \leftarrow\ [|\ \alpha\,X_2\,\textrm{sym}(X_1) + \beta Y\ |]\ \mathbf{in\ } e
    \Rightarrow \\
    & \mathbf{let\ } ((X_1,X_2),Y)\ = \mathbf{symm\ } \mathbf{true}
        \ \alpha\ \_ \ X_1 \_\ X_2\ \beta\ Y\ \mathbf{in\ } e
\\[0.1\baselineskip]
%gemm
    \mathbf{let\ } &Y \leftarrow [|\ \alpha X_1^{T?} X_2^{T?} + \beta Y\ |]\ \mathbf{in\ } e
    \Rightarrow \\
    & \mathbf{let\ } ((X_1,X_2),Y)\ =\ \mathbf{gemm\ } \alpha
        \ \_\ \left(X_1, \substack{\mathbf{true}\\\mathbf{false}}\right)
        \ \_\ \left(X_2, \substack{\mathbf{true}\\\mathbf{false}}\right)
        \ \beta\ Y\ \mathbf{in\ } e
\end{align*}

\caption{Purely syntactic pattern-matching translations of
    matrix expressions.}\label{fig:lang_matexp}
\end{figure}

\subsubsection{Code Generation} is a straightforward mapping from \lang's core
constructs to high-level OCaml ones. We embed \lang's type- and term-
constructors into OCaml as a sanity check on the output
(Figure~\ref{fig:type_grammar}).

\begin{figure}[t]
    \centering
    \begin{minipage}{.3\textwidth}
        \centering
        \begin{grammar}
            <f> ::= `'
            \alt <fc>
            \alt `Z'
            \alt `S' <f>

            <t> ::= `'
            \alt `unit'
            \alt `bool'
            \alt `int'
            \alt `elt'
            \alt <f> `arr'
            \alt <f> `mat'
            \alt `!' <t>
            \alt \synt{$\forall$} <fc.> <t>
            \alt <t> \lit{$\otimes$} \synt{$t'$}
            \alt <t> \lit{$\multimap$} \synt{$t'$}
        \end{grammar}
    \end{minipage}
    \begin{minipage}{.3\textwidth}
        \centering
        \begin{minted}[fontsize=\small]{ocaml}
module Arr =
  Owl.Dense.Ndarray.D

@@

type 'a bang =
  Many of 'a
  [@@unboxed]
        \end{minted}
    \end{minipage}
    \begin{minipage}{.3\textwidth}
        \begin{align*}
            [\![ f\!c ]\!] &= \texttt{'fc} \\
            [\![ \textbf{Z} ]\!] &= \texttt{z}\\
            [\![ \textbf{S} \, f ]\!] &= [\![ f ]\!]\, \texttt{s}\\
            [\![ \textbf{unit} ]\!] &= \texttt{unit}\\
            [\![ \textbf{bool} ]\!] &= \texttt{bool}\\
            [\![ \textbf{int} ]\!] &= \texttt{int}\\
            [\![ \textbf{elt} ]\!] &= \texttt{float}\\
            [\![ f\, \textbf{arr} ]\!] &= [\![ f ]\!]\, \texttt{arr}\\
            [\![ f\, \textbf{mat} ]\!] &= [\![ f ]\!]\, \texttt{mat}\\
            [\![ \textbf{!} \, t ]\!] &= [\![ t ]\!]\, \texttt{bang}\\
            [\![ \forall f\!c.\, t ]\!] &= [\![ t ]\!]\\
            [\![ t \otimes t' ]\!] &= [\![ t ]\!] \texttt{*} [\![ t' ]\!]\\
            [\![ t \multimap t' ]\!] &= [\![ t ]\!] \rightarrow [\![ t' ]\!]
        \end{align*}
    \end{minipage}
    \caption{\lang's type grammar (left) and its embedding into OCaml
        (right).}\label{fig:type_grammar}
\end{figure}

>>>>

As seen earlier (Figure~\ref{fig:lang_desugar}), desugaring is conventional.
Matrix expressions are translated into BLAS/LAPACK calls via purely syntactic
pattern-matching (also seen earlier in Figure~\ref{fig:lang_matexp}).

@@

\subsubsection{Code Generation}

Code generation is a straightforward mapping from \lang's core constructs to
high-level OCaml ones. We embed \lang's type- and term- constructors into OCaml
as a sanity check on the output (Figure~\ref{fig:type_grammar}).

\begin{figure}[t]
    \centering
    \begin{minipage}{.3\textwidth}
        \centering
        \begin{grammar}
            <f> ::= `'
            \alt <'fc>
            \alt `z'
            \alt <f> `s'

            <t> ::= `'
            \alt `unit'
            \alt `bool'
            \alt `int'
            \alt `elt'
            \alt <f> `arr'
            \alt <f> `mat'
            \alt `!' <t>
            \alt <'fc.> <t>
            \alt <t> \lit{$\otimes$} \synt{$t'$}
            \alt <t> \lit{$\multimap$} \synt{$t'$}
        \end{grammar}
    \end{minipage}
    \begin{minipage}{.3\textwidth}
        \centering
        \begin{minted}[fontsize=\small]{ocaml}
module Arr =
  Owl.Dense.Ndarray.D

@@

type 'a bang =
  Many of 'a
  [@@unboxed]
        \end{minted}
    \end{minipage}
    \begin{minipage}{.3\textwidth}
        \begin{align*}
            [\![ '\!f\!c ]\!] &= \texttt{'fc} \\
            [\![ \textbf{z} ]\!] &= \texttt{z}\\
            [\![ f \, \textbf{s} ]\!] &= [\![ f ]\!]\, \texttt{s}\\
            [\![ \textbf{unit} ]\!] &= \texttt{unit}\\
            [\![ \textbf{bool} ]\!] &= \texttt{bool}\\
            [\![ \textbf{int} ]\!] &= \texttt{int}\\
            [\![ \textbf{elt} ]\!] &= \texttt{float}\\
            [\![ f\, \textbf{arr} ]\!] &= [\![ f ]\!]\, \texttt{arr}\\
            [\![ f\, \textbf{mat} ]\!] &= [\![ f ]\!]\, \texttt{mat}\\
            [\![ \textbf{!} \, t ]\!] &= [\![ t ]\!]\, \texttt{bang}\\
            [\![  '\!f\!c.\, t ]\!] &= [\![ t ]\!]\\
            [\![ t \otimes t' ]\!] &= [\![ t ]\!] \texttt{*} [\![ t' ]\!]\\
            [\![ t \multimap t' ]\!] &= [\![ t ]\!] \rightarrow [\![ t' ]\!]
        \end{align*}
    \end{minipage}
    \caption{\lang's type grammar (left) and its embedding into OCaml
        (right).}\label{fig:type_grammar}
\end{figure}
